NCAM Explorer

### 2020-03-15 analysis/paper/response_letter.Rmd

Daniel Goethel, PhD  
Southeast Fisheries Science Center  
National Marine Fisheries Service  
National Oceanic and Atmospheric Administration  
75 Virginia Beach Drive  
Miami, FL 33133  
USA

`r format(Sys.time(), '%Y-%m-%d')`

Dear Dr. Goethel,

Thank you for considering another revision of manuscript cjfas-2019-0424,  **"Improving the communication and accessibility of stock assessment using interactive visualization tools"** by Paul M. Regular, Gregory J. Robertson, Robert Rogers and Keith P. Lewis. We are very grateful for the constructive and detailed reviews provided and we have made every effort to do justice to the reviewer's suggestions. In the process, we believe our manuscript has been greatly improved. Most notably we have expanded our introduction and discussion to include a more through exploration of the literature, emphasizing advances in fisheries science. We have also clarified the intent and audience for the tools highlighted in our manuscript and these changes should curtail impressions that these tools are supposed to be stand-alone documents. These changes should also alleviate impressions that interactive tools serve as a complete replacement for static documents when, in fact, they are best used to complement static documentation. Please see below for more details on the changes we made in response to the reviews.

@@

*We have gone through our manuscript and we have tempered several conclusions. Some remain as our perspective is based on extensive verbal feedback at meetings from stakeholders and following tutorials the primary author has provided on this topic. In the "Towards open stock assessment" section, we also clarify why we think these interactive tools improve accessibility.*

*We acknowledge that a formal survey would be more scientific and it would offer insights into ways of improving the dashboards, however, such a formal approach was outside the scope of this work. Likewise, a jurisdictional survey to learn about and speak to experiences outside our region would be an excellent addition to the paper, however, this too was outside the scope of this work. If we implemented such surveys we would likely be submitting this paper as a research article with stronger categorical statements.*

Similarly, there are a number of discussion points raised by the reviewers that should be incorporated into the manuscript. For instance, reviewer 3 suggests discussion on how to incorporate graphics that compare results across multiple model runs, how to streamline versioning and code updates, how to ensure compatibility among graphics code and assessment outputs as either are updated or reparametrized, and what the benefits/difficulties are of developing generic versus study-specific visualization tools.

*These are important points, several of which the first author coincidentally discussed with colleagues while this paper was in review. We have added a new section, "From bespoke to general tools", where we discuss the challenges and potential benefits of developing generic tools. We also acknowledge that the tools can be modified to visualize multiple model runs. In the following section, "Towards open stock assessment", we note that versioning can be traced using GitHub. The only item we did not cover was ensuring compatibility; this is where automated testing via TravisCI and the testthat package are sure to help, however, we did not add this to the discussion because 1) these are programming tools the authors have limited experience with, and 2) moving into technical areas beyond the scope of this paper.*

@@

I also agree that the link between interactive and static reports or presentations is unclear. It would seem that there needs to be a mix of approaches, but this is not really discussed. Similarly, what tips would the authors give for presenting results using interactive graphics? One element that is often useful with static presentations is the building of a ‘story or progress line’ (especially in model building or continuity exercises), which might be difficult to replicate if only using interactive graphics. It does not seem like an all or nothing approach is warranted, but that is the impression that I received from the article. I would suggest carefully reworking the discussion of how interactive graphics can be used when presenting to stakeholders and how static and interactive methods can be used synergistically.

*This is a very good point and, in hindsight, we see why the paper leaves this impression. To try and rectify this, we have clarified the intent of interactive tools in several places in the paper. We now explicitly note in the introduction that these tools are not meant to replace static modes of communication, rather they are often intended to supplement standard documents. This point is reiterated in the  "Towards open stock assessment" section, where we also highlight an example from Appendix A to make the point that advisory documents (i.e. a linear story) can be interactive. As with the previous iteration, we contend that a non-linear format can be useful for evaluating a stock assessment model. That being said, we take your point that a story may be more suitable in some cases. This format is possible with interactive documents as well because dashboards can be nested within html slide-shows, or dashboards can include "story-boards". These are very flexible tools and in this iteration of the manuscript we hope we have clarified their flexibility without discounting static or linear formats. In short, we want to make the point that interactive tools are a potentially powerful tool to add to a biologists’ toolkit.*

@@

*See "stand alone" comment above.*

Partial catches:

-Need explanation as well as context, the tool overall should be able to tell a story by itself but it feels like a somewhat chaotic assembly of cut out figures from a stock assessment report.

*See "stand alone" comment above. Nevertheless, we have added equations and an additional figure to provide more context. We have also added more text and equations to subsequent tabs.*

@@

*See "stand alone" comment above.*

References:

-Should be more references and they should be linked to the documents.

*More references have been added (all of the key references for the accepted assessment model).*

@@

Catch tab:

-Everywhere in the tool there is insufficient labelling and cryptic legends, for example it is not clear what is meant by ‘bounds’ or what is represented by the shaded area -the choices of colour and line types are questionable and inconsistent throughout the tool -there are two shaded areas in the bottom graph (absolute landings) without a key

*Agreed, the colors were not ideal. We have made the colour scheme more consistent throughout the tool.*

@@

On the right side:

-there should be consistency in how observations and estimation are signified, going from a dashed line to solid circles adds unnecessary confusion -the graphs can have better titles such as ‘Age 2’ rather than ‘2’, for instance -other graphs, such as ‘fits by age’ are unwieldly and confusing, this, like a majority of graphs, does not represent a good use of data visualisation principles and betrays the authors’ lack of experience with graphic design. When there is a lot of data in one graph, making sure that interactivity works well is paramount, currently selecting a specific line is not so clear. Use different colour scheme.

*True, the fits by age plots were confusing and only useful if the user selected a specific age using the legend. These plots have been removed and replaced by plots with a slider for selecting specific ages. Now the facet plot allows a user to assess the fits at a glance with one plot and, if interested in the details, shift to the frame plot to take a closer look at the fits to specific ages.*

@@

Overall there are too many graphs, offered in such a way as it is difficult to prioritise information. The design can be far more ergonomic, with graphs organised more helpfully and with a sense of hierarchy and a narrative, encouraging to explore key results first, while offering extra levels of details as a choice to those who want to pursue a particular matter further. There as an inconsistent and suboptimal from a design point of view use of colour. A data visualisation professional should be able to help with making a good use of colour to encode information and to help the user navigate through results without getting overwhelmed. I would particularly recommend enabling the user to make side by side comparisons of different combination of graphs that are currently in fixed/distinct places in the tool.

*Undoubtedly, there are many ways to organize this information and, given this comment, we have tried to improve the layout of the tool. Overall, we have done our best to organize it in a way that makes sense to the aforementioned experts at the stock assessment meeting. That is, the pages follow the general order in which our stock assessments tend to flow: 1) assess fits and residuals, 2) assess retrospective plots and derived quantities, 3) if 1 and 2 pass peer review, assess the projections.*

*We know the tool is far from perfect and agree that it would benefit from input from programmers and/or graphic design experts but, unfortunately, we lack funding to pursue such consultations. This is a reality that many biologists face and a key point of our paper is that this limitation should not preclude the use of these tools.*

@@

This is a clearly written and organized manuscript that presents an application of novel, open-source tools to disseminate stock assessment inputs and results. The interactive visualization tools were developed by others but modified by the authors to suit their needs; the original developers are credited appropriately. The authors make a strong case for using these tools for visualization of stock assessment results, and in that way, this paper is fitting for a perspective since it promotes a new paradigm for dissemination of stock assessments. There are no new research results, nor is there anything controversial in this piece. An important and laudable message of the paper is that these tools can help the transition to "a culture of open science and reproducible research."

*We thank the Reviewer for the kind comments.*

The manuscript repeatedly mentions the usefulness of these tools for stakeholders, both for data exploration and data modeling. In the data exploration figures (Figures 1 through 3), the interfaces would be more digestible to stakeholders if they included an annotation of what the user is viewing. For example, the Figure 1 view could include a note specifying that the error bars represent 95% confidence intervals, and species names should be used rather than, or in addition to, codes. Figure 2 could be improved by including units for the length bins, and Figure 3 should have a legend for the kernals. Each could also have a text field that gives an overview of the what the figure is showing and how a user can interact with the figure.

*All excellent suggestions, which will improve future iterations of the tools. These two tools, unlike the NCAM dashboard, were mainly built for use by the developer themselves. When these tools move to being used by larger communities, further documentation will be essential. Our intent in showing these two ‘in-house’ tools was to demonstrate that interactive visualization tools can be a valuable addition to an analysts tool box.*

@@

More broadly, the NCAM Explorer (Supplement 2) would also be more functional with increased annotation. The background tab provides important context but should be expanded beyond the "terse point-form" to serve as a user guide for each tab. These improvements apply to the tools themselves rather than to the manuscript.

*We agree with these comments and have added text to the Background tab and added references for background.  See "stand alone" comment above.  We have tried to strike a balance between addressing the spirit of the Reviewers comments and unnecessarily cluttering the dashboard.*

The one area where I think the manuscript should be expanded is in the "Towards open stock-assessment" section. The main reason given for increasing the transparency of stock assessment through tools such as these is to improve peer review. However, as mentioned at several other points in the manuscript, these tools are also valuable for increasing stakeholder engagement in and understanding of stock assessment. That function warrants equal attention to peer review in this section and could present ideas for how to introduce these tools to stakeholders and how to help them sift through the large volume of data to find what is most relevant to them.

*We have expanded the "Torwards open stock-assessment" section and, in the process, we hope we have clarified how these tools improve accessability and transparency by making it easy for others to access the details.*

@@

Lines 262: The statement "Although these tools do not reveal the entire data pipeline" may be contradicted by the ICES Transparent Assessment Framework (https://taf.ices.dk/app/about) which archives the full process of data processing to running assessment models in a reproducible way and should be added to Appendix A.

*We thank the Reviewer for drawing our attention to the TAF. We have now re-structured our "Towards open stock assessment" section to highlight TAF and several other open science initiatives in fisheries research.*

Figure 2: I realize this figure is only for illustrative purposes, but it might be good to add a note about the strange pattern in the length compositions of American plaice in the 1980s. The spiky pattern with peaks in the same bins each year seems like an artifact of rounding error or conversion from some other type of length measurement.

*This is a very astute observation and it is most likely related to changes in how Plaice length frequencies were recorded (3 to 1 cm scale). While this is an important detail to be cognisant of when analyzing these data, we have opted not to get into the minutiae of the sampling as it is outside the scope of this paper.*

### 2020-04-17 analysis/paper/manuscript.Rmd
Revisions following latest CJFAS review

Quantitative stock assessment plays a central role in modern fisheries management [@hilborn1992]. Over time, and as new methods are developed, there is an increasing amount of data available to inform stock assessments. These data might be richer information on stock structure using an increasing array of markers and biomarkers, and/or the continued lengthening of population status and catch time series. Concurrently, analytical methods are now able to integrate many data sources into one stock assessment model [@maunder2013]. While these advances have surely improved the advice provided at stock assessments, it has become more difficult to communicate the basis of this advice in a meaningful way. Visualizations play a critical role in communicating this information to stakeholders [@levontin2017], however, traditional formats (e.g. slides) are often insufficient to convey the richness of information available and the static and sequential nature of these formats can stifle meaningful discussions. A solution to this problem is the effective use of interactive visualization tools [@keena2016]. These are common tools we use every day on a range of web sites, but their use is no longer restricted to web site developers as these tools are being integrated into software commonly used by the research community [@perkel2018]. As a result, an increasing number of scientists across a wide range of disciplines are starting to apply interactive visualization tools to explore and communicate their results [e.g. @jones2016; @yeatman2018; @letcher2018].

@@

We first demonstrate the application of interactive tools to data that are commonly presented in stock assessments to show how they can simplify the detailed exploration of data from long-term monitoring programs. Second, we focus on the modeling aspect of stock assessment and demonstrate how interactive tools can be used to explore, diagnose and communicate results from an integrated assessment model. This structure corresponds to two important  steps in the stock assessment process: 1) data exploration, and 2) data modeling. In both cases interactive visuals are nested inside dynamic documents called "dashboards". The concept of using a dashboard was borrowed from the business community where they are frequently used to group a series of interactive visuals and tables to provide at-a-glance views of key performance indicators. These dashboards were constructed within the framework of the R programming language [@R] and the RStudio IDE [@Rstudio] using three or more of the packages listed in Table \@ref(tab:pkg). See Supplement 1 for a "getting started" tutorial that highlights one approach to constructing a dashboard using R; using this as a guide, an R user can start constructing basic dashboards within a day.

@@

Northern cod (NAFO Divisions 2J3KL) has a rich history of tagging, starting in in 1954 [@taggart1995] and continuing to this day. The tagging and recovery data are captured in a standardized database, with fields typical of most tagging programs. This database has over 600,000 records as of early 2019, with 2,000-10,000 tags deployed annually in recent decades. The tagging and capture data are used in the current assessment model for this stock [@cadigan2016], but tools to explore this extensive data set were limited, especially from a spatial perspective. To begin to explore and understand this large data set, we built a simple application using shiny and leaflet (Table \@ref(tab:pkg)) that allows a user to quickly and dynamically subset the data (e.g. ranges of release and capture years, specific geographic locations). As the application developed, it was incorporated into shinydashboard, to take advantage of the number of layout options available in shinydashboard (Table \@ref(tab:pkg)). Given the large number of tags released, often at nearby sites, visualizing the data with static mapping was particularly challenging. The markercluster (https://github.com/Leaflet/Leaflet.markercluster) function available in leaflet was particularly useful as a means to dynamically scale the level of pooling of spatial points (Figure \@ref(fig:map)). This basic mapping tool allowed us to quickly become familiar with the data, identify outliers and incorrect data entries, as well as explore options on how to spatially pool the data for subsequent demographic analysis. Further tabs were added to provide basic summaries of the selected data (Figure \@ref(fig:summary)).

```{r map, echo = FALSE, fig.cap = 'Screen shot of cod tag mapping tool using shinydashboard and leaflet. The markercluster function dynamically splits or pools tagging locations (red, orange, green or yellow points) depending on zoom level, the recoveries positions (blue) are much fewer, and are left to be plotted individually at all scales. Options to include pop up labels are included, so specific information on each point can be retrieved with a mouse click (in this case: tag number, fish length, date released, and date captured), which is particularly useful when error checking.'}
knitr::include_graphics("figures/tagging_dashboard_map_tab.png")
```

```{r summary, echo = FALSE, fig.cap = 'Basic summaries of the recovery data from the tags selected within the shiny dashboard cod tag mapping tool. In this case, histograms and summary statistics of the recovery positions are returned, along with a simple map of kernels showing the 2D spatial distribution of the selected tag recoveries.'}
knitr::include_graphics("figures/tagging_dashboard_summary_tab.png")
```

@@

Synthesizing data from multiple sources presents a key challenge to stock assessment. Analyses of different data sources were traditionally carried out independently and the summaries or parameters from these analyses were used in the assessment model. This approach, however, is less than ideal because information may be lost and uncertainty may be unaccounted for when we "do statistics on the statistics" [@link1999; @maunder2013]. Such issues have largely been curtailed in contemporary stock assessments thanks to advances in statistical computing that have facilitated the analysis of all available data, in as raw a form as appropriate, in a single integrated analysis [@maunder2013]. Specifically, statistical modeling tools such as JAGS [@plummer2003], AD Model Builder [@fournier2012] and Template Model Builder [@kristensen2015] allow the construction of a joint likelihood for an array of observations to, in theory, extract as much information as possible about the biological and fishery processes. From a computational perspective, analyses of a variety of large data sets has never been easier. However, from a human perspective, contemporary stock assessment biologists are faced with the challenge of understanding and integrating data from multiple sources into a single model and communicating the methods and results to stakeholders and fisheries managers. This challenge was palpable for the data-rich case of Northern cod.

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions and partial fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model simplifies thousands of historical data points into quantities such as recruitment, spawning stock biomass, fishing mortality and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model output. The traditional approach would involve producing static presentations and documents with a series of figures and tables.  However, with large amounts of model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and the stakeholders involved for at least two reasons. First, it is no longer feasible for the analyst to include and describe every figure and table produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

In the pursuit of an easier and more efficient way to communicate results from NCAM, an interactive "dashboard", called NCAM explorer, was developed for the 2018 assessment of Northern cod [@dwyer2019]. We used R-based packages (Table \@ref(tab:pkg)) to construct a tool for exploring the input and output of NCAM, specifically the flexdashboard package to group interactive plotly-based visuals into a dynamic document. We also used the crosstalk package to link the data displayed across multiple plots. Via R Markdown, the dashboard is rendered into a self-contained html file that is reproducible, interactive, and easy to update following modifications to the model or the addition of new data. The intent of this tool was not to replace detailed documentation on the model [i.e. @cadigan2016; @dwyer2019], however, it was built to serve as an interactive alternative to a standard presentation for communicating the latest results to an audience with a working knowledge of the stock and model. Explanatory text and detail were therefore kept to a minimum, allowing the audience to explore the figures and data unencumbered by verbose descriptions.

The NCAM dashboard (Supplement 2) contains a series of pages, the first of which provides terse point-form background on the model (page named "Background"). Subsequent pages provide a series of diagnostic plots for assessing model fits to catch ("Landings"; accessed from the "Fishery" drop-down menu), survey ("RV survey", "SN survey" and "SS survey"; accessed from the "Surveys" drop-down menu), and tagging ("Tagging") data. For instance, the "RV survey" page includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes pages focused on model estimates such as selectivity ("Selectivity"; accessed from the "Fishery" drop-down menu) and catchability ("Catchability"; accessed from the "Surveys" drop-down menu). Pages for viewing trends in various population parameters are included under the "Trends" drop-down menu, where the "Retro" page displays results from a retrospective analysis, the "Comps" page displays population trends from different models, the "Status and rates" page displays trends in stock size and vital rates (Figure \@ref(fig:trends)), and the "Productivity" page focuses on trends in SSB and recruitment. Finally, details on the projections are accessed from the "Projections" drop-down menu ("Assumptions", "Past projections", "Retro projections", and "Results" pages), and key inputs and outputs are accessed from "Tables" drop-down menu ("Inputs", "Settings", "Outputs" pages). The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents. However, there are two key benefits of this approach over the *modus operandi* of producing static documents and slides.  First, interactive plots nested in a dashboard permit relatively easy and efficient access to the details as it replaces scrolling through tens, if not hundreds, of pages or slides with mouse-clicks across pages holding data-rich illustrations (i.e. both broader patterns and finer details in the data are accessible via zooming and tooltips).  Second, the automated nature of the dashboard circumvents the monotonous, time-consuming and error-prone task of copying and pasting figures, tables and values into documents and slides, although, we note that this point is not exclusive to the proposed approach (e.g. slide shows can easily be made using R markdown). Both benefits expedite the process of exploring a range of model configurations as the automated output facilitate quick views of standard diagnostics and the interactive plots facilitate detailed explorations and comparisons of models with different configurations. For instance, explorations of the residual plots in the dashboard allowed the authors to quickly identify conflicts between two surveys used in the model. Potential solutions to this problem were then explored by modifying the configuration of the model and these changes were quickly evaluated using a dashboard that is produced in seconds.

The dashboard also makes it easy to share the results with colleagues and stakeholders as it is rendered into a self-contained html file. This allows others to independently scrutinize details of both data and the model that are typically only accessible to the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model. For instance, visualizations depicting the model assumptions, process errors and confidence intervals around the projections raised important questions on the impact of the assumptions on the projections from the model. These questions were raised during the first assessment meeting in which this tool was used and the ensuing discussions left us with the impression that this tool helped the assessment biologists, academics, resource managers and fisheries representatives in the room gain a deeper understanding of the uncertainties and the patterns in the risk tables. Feedback and personal testimony from multiple stakeholders who have used the dashboard reaffirmed our impression that the tool is helpful as there was general agreement that access to the details eases communication and relieves any potential concerns that something is hidden.

```{r rv, echo = FALSE, fig.cap = 'Screenshot of the "RV survey" tab from the NCAM dashboard where total observed (dots) and NCAM model predicted values (lines) for the DFO RV survey index are shown in the upper left panel and scaled matrix plot of age-disaggregated standardized log residuals are shown in the lower left panel (blue = positive, red = negative, symbols scaled by size; grey = index values of zero). Age-disaggregated observed (dots) and predicted (lines) values from the DFO RV survey are shown in the right panel.'}
knitr::include_graphics("figures/NCAM_dashboard_rv-survey_tab.png")
```

```{r trends, echo = FALSE, fig.cap = 'Screenshot of the "Trends" tab which displays estimates of recruitment, stock size, and stock size relative to B~lim~ (left panels) and mortality rates (F, M, Z, right panels).'}
knitr::include_graphics("figures/NCAM_dashboard_trends_tab.png")
```

@@

The interactive tools highlighted here were custom built to solve data visualization challenges that are common to many fisheries management organizations but the solutions presented are specific to the Northwest Atlantic Fisheries Centre. The concept is therefore broadly applicable, however, the underlying code will require modifications to accommodate region and stock specific data management and modelling approaches. Even within our own region, different approaches and needs meant that the code behind NCAM explorer (Supplement 2) required modifications when applied to a model developed for another cod stock. Nevertheless, considerable chunks of code were recycled and, in the process, new features were added to allow the visualization of multiple model configurations within the one dashboard. Additionally, a French colleague is currently modifying the tool to visualize output from SAM, a state-space stock assessment model widely used to assess ICES stocks [@nielsen2014]. Likewise, the concept behind RStrap Explorer and the tagging tool could be applied to other regions and the most efficient approach may be to build upon existing general frameworks [e.g. StoX software; @johnsen2019]. These are, of course, early days in the development and application of interactive tools to stock assessment. Because the tools we highlight are themselves general, flexible and accessible, we suspect that bespoke applications will continue to be developed in parallel with more general stock assessment dashboards depending on user needs and preferences. Regardless of the approach, a common goal is to make the workflow more efficient, reproducible and transparent.

>>>>

Quantitative stock assessment plays a central role in modern fisheries management [@hilborn1992]. Over time, and as new methods are developed, there is an increasing amount of data available to inform stock assessments. These data might be richer information on stock structure using an increasing array of markers and biomarkers, and/or the continued lengthening of population status and catch time series. Concurrently, analytical methods are now able to integrate many data sources into one stock assessment model [@maunder2013]. While these advances have surely improved the advice provided on stock status, it has become more difficult to communicate the basis of this advice in a meaningful way. Visualizations play a critical role in communicating this information to stakeholders [@levontin2017]. However, traditional formats (e.g. slides) are often insufficient to convey the richness of information available and the static and sequential nature of these formats can stifle meaningful discussions. A solution to this problem is the effective use of interactive visualization tools [@keena2016]. These are common tools we use every day on a range of web sites, but their use is no longer restricted to web site developers as these tools are being integrated into software commonly used by the research community [@perkel2018]. As a result, an increasing number of scientists across a wide range of disciplines are starting to apply interactive visualization tools to explore and communicate their results [e.g. @jones2016; @yeatman2018; @letcher2018].

@@

We first demonstrate the application of interactive tools to data that are commonly presented in stock assessments to show how they can simplify the detailed exploration of data from long-term monitoring programs. Second, we focus on the modeling aspect of stock assessment and demonstrate how interactive tools can be used to explore, diagnose, and communicate results from an integrated assessment model. This structure corresponds to two important  steps in the stock assessment process: 1) data exploration, and 2) data modeling. In both cases interactive visuals are nested inside dynamic documents called "dashboards". The concept of using a dashboard was borrowed from the business community where they are frequently used to group a series of interactive visuals and tables to provide at-a-glance views of key performance indicators. These dashboards were constructed within the framework of the R programming language [@R] and the RStudio IDE [@Rstudio] using three or more of the packages listed in Table \@ref(tab:pkg). See Supplement 1 for a "getting started" tutorial that highlights one approach to constructing a dashboard using R; using this as a guide, an R user can start constructing basic dashboards within a day.

@@

Northern cod (NAFO Divisions 2J3KL) has a rich history of tagging, starting as a dedicated program in 1954 [@taggart1995] and continuing to this day. The tagging and recovery data are captured in a standardized database, with fields typical of most tagging programs. This database has over 600,000 records as of early 2019, with 2,000-10,000 tags deployed annually in recent decades. The tagging and recovery data are used in the current assessment model for this stock [@cadigan2016], but tools to explore this extensive data set were limited, especially from a spatial perspective. To begin to explore and understand this large data set, we built a simple application using shiny and leaflet (Table \@ref(tab:pkg)) that allows a user to quickly and dynamically subset the data (e.g. ranges of tag release and tag recovery years, specific geographic locations). As the application developed, it was incorporated into shinydashboard, to take advantage of the number of layout options available in shinydashboard (Table \@ref(tab:pkg)). Given the large number of tags released, often at nearby sites, visualizing the data with static mapping was particularly challenging. The markercluster (https://github.com/Leaflet/Leaflet.markercluster) function available in leaflet was particularly useful as a means to dynamically scale the level of pooling of spatial points (Figure \@ref(fig:map)). This basic mapping tool allowed us to quickly become familiar with the data, identify outliers and incorrect data entries, as well as explore options on how to spatially pool the data for subsequent demographic analysis. Further tabs were added to provide basic summaries of the selected data (Figure \@ref(fig:summary)).

```{r map, echo = FALSE, fig.cap = 'A screen shot of cod tag mapping tool using shinydashboard and leaflet. The markercluster function dynamically splits or pools tagging locations (red, orange, green or yellow points) depending on zoom level, the recovery positions (blue) are much fewer, and are left to be plotted individually at all scales. Options to include pop up labels are included, so specific information on each point can be retrieved with a mouse click (in this case: tag number, fish length, date released, and date recovered), which is particularly useful when error checking.'}
knitr::include_graphics("figures/tagging_dashboard_map_tab.png")
```

```{r summary, echo = FALSE, fig.cap = 'Basic summaries of the recovery data from the tags selected within the shiny dashboard cod tag mapping tool. In this case, histograms and summary statistics of the recovery positions are returned, along with a simple map of kernels showing the spatial distribution of the selected tag recoveries.'}
knitr::include_graphics("figures/tagging_dashboard_summary_tab.png")
```

@@

Synthesizing data from multiple sources presents a key challenge to stock assessment. Analyses of different data sources were traditionally carried out independently and the summaries or parameters from these analyses were used in the assessment model. This approach, however, is less than ideal because information may be lost and uncertainty may be unaccounted for when we "do statistics on the statistics" [@link1999; @maunder2013]. ). Such issues have largely been curtailed in contemporary stock assessments as advances in statistical computing have enabled integrated analyses of data from multiple monitoring programs [@maunder2013]. Specifically, statistical modeling tools such as JAGS [@plummer2003], AD Model Builder [@fournier2012], and Template Model Builder [TMB; @kristensen2015] allow the construction of a joint likelihood for an array of observations to, in theory, extract as much information as possible about the biological and fishery processes. From a computational perspective, analyses of a variety of large data sets has never been easier. However, from a human perspective, contemporary stock assessment biologists are faced with the challenge of understanding and integrating data from multiple sources into a single model and communicating the methods and results to stakeholders and fisheries managers. This challenge was palpable for the data-rich case of Northern cod.

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions, and reported fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model simplifies thousands of data points into quantities such as annual recruitment, spawning stock biomass, fishing mortality, and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model output. The traditional approach would involve producing static presentations and documents with a series of figures and tables.  However, with numerous model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and the stakeholders involved for at least two reasons. First, it is no longer feasible for the analyst to include and describe every figure and table produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

In the pursuit of an easier and more efficient way to communicate results from NCAM, an interactive "dashboard", called NCAM explorer, was developed for the 2018 assessment of Northern cod [@dwyer2019]. We used R-based packages (Table \@ref(tab:pkg)) to construct a tool for exploring the input and output of NCAM, specifically the flexdashboard package to group interactive plotly-based visuals into a dynamic document. We also used the crosstalk package to link the data displayed across multiple plots. Via R Markdown, the dashboard is rendered into a self-contained html file that is reproducible, interactive, and easy to update following modifications to the model or the addition of new data. The intent of this tool was not to replace detailed documentation on the model [i.e. @cadigan2016; @dwyer2019], however, it was built to serve as an interactive alternative to a standard presentation for communicating the latest results to an audience with a working knowledge of the stock and assessment model. Explanatory text and detail were therefore kept to a minimum, allowing the audience to explore the figures and data unencumbered by verbose descriptions.

The NCAM dashboard (Supplement 2) contains a series of pages, the first of which provides terse point-form background on the model (page named "Background"). Subsequent pages provide a series of diagnostic plots for assessing model fits to catch ("Landings"; accessed from the "Fishery" drop-down menu), survey ("RV survey", "SN survey" and "SS survey"; accessed from the "Surveys" drop-down menu), and tagging ("Tagging") data. For instance, the "RV survey" page includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes pages focused on model estimates such as selectivity ("Selectivity"; accessed from the "Fishery" drop-down menu) and catchability ("Catchability"; accessed from the "Surveys" drop-down menu). Pages for viewing trends in various population parameters are included under the "Trends" drop-down menu, where the "Retrospective analysis" page displays results from a retrospective analysis, the "Model comparison" page displays population trends from different model runs, the "Status and rates" page displays trends in stock size and vital rates (Figure \@ref(fig:trends)), and the "Productivity" page focuses on trends in Spawning Stock Biomass (SSB) and recruitment. Finally, details on the projections are accessed from the "Projections" drop-down menu ("Assumptions", "Past projections", "Retro projections", and "Results" pages), and key inputs and outputs are accessed from "Tables" drop-down menu ("Inputs", "Settings", "Outputs" pages). The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents. However, there are two key benefits of this approach over the *modus operandi* of producing static documents and slides.  First, interactive plots nested in a dashboard permit relatively easy and efficient access to the details as it replaces scrolling through tens, if not hundreds, of pages or slides with mouse-clicks across a smaller number of dynamic pages holding data-rich illustrations (i.e. both broader patterns and finer details in the data are accessible via zooming and tooltips).  Second, the automated nature of the dashboard circumvents the monotonous, time-consuming and error-prone task of copying and pasting figures, tables, and values into documents and slides, although, we note that this issue can be addressed through other approaches (e.g. slide shows can easily be made using R markdown). Both benefits expedite the process of exploring a range of model configurations as the automated output facilitate quick views of standard diagnostics and the interactive plots facilitate detailed explorations and comparisons of models with different configurations. For instance, explorations of the residual plots in the dashboard allowed the authors to quickly identify conflicts between two surveys used in the model. Potential solutions to this problem were then explored by modifying the configuration of the model and these changes were quickly evaluated using a dashboard that is produced in seconds.

The dashboard also makes it easy to share the results with colleagues and stakeholders as it is rendered into a self-contained html file. This allows others to independently scrutinize details of both data and the model that are typically only accessible to the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model. For instance, visualizations depicting the model assumptions, process errors and confidence intervals around the projections raised important questions on the impact of the assumptions on the projections from the model. These questions were raised during the first assessment meeting in which this tool was used and the ensuing discussions left us with the impression that this tool helped the assessment biologists, academics, resource managers, and fisheries representatives in the room gain a deeper understanding of the uncertainties and the patterns in the risk tables. Feedback and personal testimony from multiple stakeholders who have used the dashboard reaffirmed our impression that the tool is helpful as there was general agreement that access to the details eases communication and relieves any potential concerns that something is hidden.

```{r rv, echo = FALSE, fig.cap = 'Screenshot of the "RV survey" tab from the NCAM dashboard where total observed (dots) and NCAM model predicted values (lines) for the DFO RV survey index are shown in the left panel and standardized log residuals by year, cohort, age, and expected value (dots), with lowess curves (lines), are shown in the right panel.'}
knitr::include_graphics("figures/NCAM_dashboard_rv-survey_tab.png")
```

```{r trends, echo = FALSE, fig.cap = 'Screenshot of the "Trends" tab which displays stock status estimates (abundance, biomass, and biomass relative to B~lim~ [average biomass through the 1980s]) on the left panel, and mortality rates (average F, M, and Z) on the right panel.'}
knitr::include_graphics("figures/NCAM_dashboard_trends_tab.png")
```

@@

The interactive tools highlighted here were custom built to solve data visualization challenges that are common to many fisheries management organizations but the solutions presented are specific to the Northwest Atlantic Fisheries Centre. The concept is therefore broadly applicable, however, the underlying code will require modifications to accommodate region and stock specific data management and modelling approaches. Even within our own region, different approaches and needs meant that the code behind NCAM explorer (Supplement 2) required modifications when applied to a model developed for another cod stock. Nevertheless, considerable chunks of code were recycled and, in the process, new features were added to allow the visualization of multiple model configurations within the one dashboard. Additionally, a French colleague is currently modifying the tool to visualize output from SAM, a state-space stock assessment model widely used to assess ICES stocks [@nielsen2014]. Likewise, the concept behind RStrap Explorer and the tagging tool could be applied to other regions and the most efficient approach may be to build upon existing general frameworks [e.g. StoX software; @johnsen2019]. These are, of course, early days in the development and application of interactive tools to stock assessment. Because the tools we highlight are themselves general, flexible, and accessible, we suspect that bespoke applications will continue to be developed in parallel with more general stock assessment dashboards depending on user needs and preferences. Regardless of the approach, a common goal is to make the workflow more efficient, reproducible, and transparent.

### 2020-04-17 inst/rmd/trends.Rmd
Revisions following latest CJFAS review

Retro {data-navmenu="Trends" `r if (is.null(fit$retro)) ".hidden"`}
================================================================================

@@

Comps {data-navmenu="Trends" `r if (is.null(tmp$comp)) ".hidden"`}
================================================================================

>>>>

Retrospective analysis {data-navmenu="Trends" `r if (is.null(fit$retro)) ".hidden"`}
================================================================================

@@

Model comparison {data-navmenu="Trends" `r if (is.null(tmp$comp)) ".hidden"`}
================================================================================
